---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# cfrnow

<!-- badges: start -->
[![R-CMD-check](https://github.com/sbfnk/cfrnow/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/sbfnk/cfrnow/actions/workflows/R-CMD-check.yaml)
[![codecov](https://codecov.io/gh/sbfnk/cfrnow/branch/main/graph/badge.svg)](https://app.codecov.io/gh/sbfnk/cfrnow)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

Real-time case fatality ratio (CFR) estimation from line-list data, using a
Bayesian mixture-cure survival model. It is the R counterpart of the CFR
component of the Julia
[`bdbv-2026-linelist-analysis`](https://github.com/sbfnk/bdbv-2026-linelist-analysis)
model, built for collaborators who work in R.

## Why

The naive `deaths / cases` ratio is biased downward in real time: recent cases
have not yet had time to die. Restricting to cases with a *resolved* outcome
(`deaths / (deaths + recoveries)`) swaps that for an upward bias, because deaths
resolve faster than recoveries, so the resolved set is enriched for deaths at any
mid-outbreak cut-off.

`cfrnow` avoids conditioning on resolution at all. Each case is fatal with
probability `cfr`; a fatal case dies at an interval-censored onset-to-death delay
`F`; every case still alive at the cut-off is right-censored, contributing the
mixture-cure survival term `1 - cfr * F(t)`, the probability that it is either
non-fatal or fatal but not yet resolved. With `F` fixed this is the
Ghani/Nishiura `deaths / sum_i F(t_i)` estimator; here `F` is co-estimated and
its uncertainty propagated. The enrichment bias is avoided because unresolved
cases are right-censored, not dropped. So cfrnow needs only death timing and who
is still a case, and does not depend on recovery dates being recorded. When they
are recorded, supplying a `recovery_delay` fits a competing-risks model that also
uses the recovery *timing* (`(1 - cfr) f_R(r)` for a recovery at `r`), which
sharpens the estimate.

Onset dates are interval-censored and, in real-time mode, `F` is right-truncated
at the cut-off. Both are handled by the analytical censored-CDF machinery
vendored from [`primarycensored`](https://primarycensored.epinowcast.org/).

## Usage

```{r usage}
library(cfrnow)

# your own onset_date / death_date line list, or a simulated one
set.seed(1)
ll <- simulate_linelist(n = 400, cfr = 0.55, onset_days = 45,
                        delay = dist.spec::Gamma(mean = 12.75, sd = 7))
d <- prepare_cfr_data(ll, obs_time = as.Date("2026-02-20"))
c(cases = d$n_cases, deaths = d$n_deaths, censored = d$n_cens)
```

You supply the onset-to-death delay as a
[dist.spec](https://epiforecasts.io/dist.spec/) distribution via the `delay`
argument (there is no default). Here the BDBV/Isiro onset-to-death prior is used:

```{r fit, eval = FALSE}
library(dist.spec)
onset_to_death <- LogNormal(meanlog = Normal(2.41, 0.2), sdlog = Normal(0.51, 0.15))
fit <- fit_cfr(d, delay = onset_to_death)
summarise_cfr(fit)
#>     quantity   mean   q2.5    q50  q97.5  rhat ess_bulk
#> 1        cfr  0.560  0.505  0.560  0.615 1.001     3100
#> 2 delay_mean 12.700 11.600 12.700 13.900 1.000     3300
#> 3   delay_sd  7.100  6.200  7.000  8.200 1.000     3000
```

Pass `obs_time = NULL` for a retrospective (fully-resolved) fit, which reduces to
the naive ratio with the delay as a nuisance.

Give a native parameter a `Normal()` prior to co-estimate it, or a fixed number /
`Fixed()` to hold it fixed; supply a `Gamma()` for a different family, or a
`recovery_delay` for the competing-risks fit:

```{r variants, eval = FALSE}
fit_cfr(d, delay = Gamma(shape = Normal(3, 1), rate = Normal(0.25, 0.1)))  # gamma
fit_cfr(d, delay = LogNormal(meanlog = 2.41, sdlog = 0.51))               # fixed-F
fit_cfr(d, delay = onset_to_death,
        recovery_delay = LogNormal(Normal(2.9, 0.3), Normal(0.5, 0.2)))   # competing risks
```

Which of these to use follows from your data: onset and death dates give the
default death-only fit; reliable recovery dates let you add a `recovery_delay`
for the competing-risks fit; and a delay you trust from elsewhere can be fixed
for the Ghani/Nishiura estimator.

You can also estimate the onset-to-death delay from a line list with
[primarycensored](https://primarycensored.epinowcast.org/)'s
`fitdistdoublecens()` (which handles the interval censoring and right-truncation)
and feed it in as the `delay` — see `vignette("cfrnow")`.

Your line list needs an `onset_date` column and a `death_date` column (`NA` for
cases that have not died). Optional columns: `onset_lower`/`onset_upper` give an
onset window that widens the primary censoring, and `recovery_date` marks a
non-fatal recovery.

`summarise_cfr()` reports `rhat`/`ess_bulk` and flags (`cfr_low_information`) when
the CFR posterior has barely moved from its prior. That is expected early in an
outbreak, when few deaths have resolved and the CFR is only weakly identified.
Treat a flagged estimate as prior-driven.

## Assumptions and caveats

The correction fixes a *timing* bias; it cannot repair the data. Read these
before quoting a number:

- **Complete death ascertainment.** A death that never reaches the line list (for
  example a community death outside a treatment centre) is silently treated as a
  survivor, biasing the CFR down. In an outbreak where ascertainment is
  ETC-centred this is the single biggest threat to validity.
- **Use the death notification date.** cfrnow counts a case as a death once its
  `death_date` is on or before the cut-off, so `death_date` should be the date
  the death entered the data, not the true date of death. If notification is
  prompt the two coincide; if it lags, the notification date lets the
  right-censoring absorb the delay and keeps `cfr` correct, whereas the true date
  of death biases the real-time `cfr` down as not-yet-notified deaths read as
  unresolved. A lag in notifying an *onset* does not bias `cfr` (it just makes
  the case absent), unless notification depends on outcome.
- **Stationary delay and homogeneous CFR.** A single onset-to-death delay and CFR
  over the whole outbreak. As treatment access (supportive care, monoclonals)
  scales up, the true CFR should fall, and a pooled estimate lags reality
  precisely when that matters. Time-varying CFR is on the roadmap below.
- **Delay family is chosen, not estimated.** Gamma and lognormal differ in tail
  weight, which affects how quickly a recent case counts as "probably cured".
  With sparse data the family is not identifiable from the line list, so check
  sensitivity by refitting with a `dist.spec::Gamma()` delay.
- **Recovery timing leans on discharge data.** The competing-risks fit is only as
  good as the recovery dates and the recovery-delay prior; where discharge
  recording is patchy, the death-only default is the safer choice.

## Roadmap

Known gaps, roughly in priority order: time-varying `cfr` (e.g. a random walk on
`logit(cfr)`); a death-reporting-delay nowcasting layer; stratification (age,
sex, vaccination); and posterior-predictive checks.

## Requirements

- [cmdstanr](https://mc-stan.org/cmdstanr/) and a working CmdStan install
- [primarycensored](https://primarycensored.epinowcast.org/)
- [dist.spec](https://epiforecasts.io/dist.spec/)

The primarycensored Stan functions are vendored into
`inst/stan/include/pcd_functions.stan`; regenerate with `data-raw/vendor_stan.R`
after upgrading primarycensored.
