---
title: "Real-time CFR with cfrnow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Real-time CFR with cfrnow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## The problem

The naive case fatality ratio, `deaths / cases`, is biased downward in real
time: recent cases have not yet had time to die. Restricting to cases with a
*resolved* outcome, `deaths / (deaths + recoveries)`, swaps that for an
upward bias, because deaths resolve faster than recoveries, so at any
mid-outbreak cut-off the resolved set is enriched for deaths.

`cfrnow` sidesteps both by never conditioning on resolution. Each case is fatal
with probability `cfr`; a fatal case dies at an interval-censored onset-to-death
delay `F`; every case still unresolved at the cut-off is right-censored,
contributing the mixture-cure survival term `1 - cfr * F(t)`. Because unresolved
cases are censored rather than dropped, the `deaths / (deaths + recoveries)`
enrichment cannot arise. Recovery dates aren't required, but when recorded a
recovered case is counted as a resolved non-fatal, which tightens the estimate.

```{r setup}
library(cfrnow)
```

## Choosing the model

Which model `fit_cfr()` runs is set by the data you have:

- With onset and death dates only, you get the default death-only fit:
  unresolved cases are censored and recoveries are left unmodelled.
- If you also have reliable recovery dates, pass a `recovery_delay` and the fit
  uses their timing (the competing-risks model).
- If you already trust an onset-to-death delay from elsewhere and the data are
  thin, fix that delay so only `cfr` is estimated (the Ghani/Nishiura estimator).

`obs_time` is a separate choice: a date gives a real-time cut-off; `NULL` gives a
retrospective fit, where every case has resolved.

## A line list

Your line list needs an `onset_date` and a `death_date` (`NA` for cases that
have not died); an optional `onset_lower`/`onset_upper` window widens the onset
censoring, and an optional `recovery_date` marks a non-fatal recovery so that
case is treated as resolved rather than censored. Here we simulate one with a
known CFR of 0.55 sampled part-way through a growing epidemic.

```{r simulate}
set.seed(1)
ll <- simulate_linelist(n = 500, cfr = 0.55, onset_days = 45,
                        delay = dist.spec::LogNormal(mean = 12.75, sd = 7))
head(ll)
```

## Preparing the data at a cut-off

`prepare_cfr_data()` classifies each case at the observation cut-off into an
observed death, a right-censored survivor, or (retrospectively) a fully-resolved
non-death.

```{r prepare}
cutoff <- max(ll$onset_date) - 3
d <- prepare_cfr_data(ll, obs_time = cutoff)
c(cases = d$n_cases, deaths = d$n_deaths, censored = d$n_cens)
naive <- d$n_deaths / d$n_cases
round(naive, 3)  # the downward-biased naive ratio
```

## Fitting

`fit_cfr()` fits the mixture-cure model with cmdstanr. You supply the
onset-to-death delay as a [dist.spec](https://epiforecasts.io/dist.spec/)
distribution (there is no default). Give a native parameter a `Normal()` prior to
co-estimate it, or a fixed number / `Fixed()` to hold it fixed (fixing the whole
delay gives the Ghani/Nishiura estimator). Here we use the BDBV/Isiro
onset-to-death prior.
This step needs a working CmdStan install and is not run when the vignette is
built.

```{r fit, eval = FALSE}
onset_to_death <- dist.spec::LogNormal(
  meanlog = dist.spec::Normal(2.41, 0.2),
  sdlog = dist.spec::Normal(0.51, 0.15)
)
fit <- fit_cfr(d, delay = onset_to_death)
summarise_cfr(fit)
#>     quantity   mean   q2.5    q50  q97.5  rhat ess_bulk
#> 1        cfr  0.579  0.528  0.579  0.625 1.001     3200
#> 2 delay_mean 12.700 11.500 12.700 13.900 1.000     3400
#> 3   delay_sd  7.200  6.300  7.200  8.200 1.000     3100
```

The corrected CFR (~0.58) recovers the true 0.55 and sits well above the naive
`r round(naive, 2)`, which is depressed by the many recent, not-yet-resolved
cases. The summary also carries `rhat`/`ess_bulk` and a `cfr_low_information`
attribute (see below). Pass `obs_time = NULL` to `prepare_cfr_data()` for a
retrospective, fully-resolved fit, which returns the naive ratio with the delay
as a nuisance.

Change the family or fix the delay by passing your own dist.spec distribution:

```{r delay-variants, eval = FALSE}
library(dist.spec)
# a gamma delay
fit_cfr(d, delay = Gamma(shape = Normal(3.3, 1), rate = Normal(0.26, 0.08)))
# a fixed delay: the Ghani/Nishiura fixed-F estimator (only cfr is estimated)
fit_cfr(d, delay = LogNormal(meanlog = 2.41, sdlog = 0.51))
```

Because the model is parameterised by the native parameters and reports
`delay_mean`/`delay_sd` as generated quantities, the summary looks the same
whichever family you choose; refit with a different family to check tail
sensitivity.

## Estimating the delay from data

Rather than hand-picking the prior, you can estimate the onset-to-death delay
from a line list with [epidist](https://epidist.epinowcast.org/), which fits the
delay with full Bayesian uncertainty while handling the double interval censoring
(onset and death each known only to the day) and, in real time, the
right-truncation. epidist works in continuous time, so pass the daily bounds of
each event and the observation time (install it from GitHub,
`epinowcast/epidist`):

```{r estimate-delay, eval = FALSE}
library(epidist)

deaths <- ll[!is.na(ll$death_date), ]
t0 <- min(deaths$onset_date)
onset <- as.numeric(deaths$onset_date - t0)
death <- as.numeric(deaths$death_date - t0)

ld <- as_epidist_linelist_data(
  onset, onset + 1,          # onset known to the day
  death, death + 1,          # death known to the day
  obs_time = max(death) + 1  # retrospective; use the real-time cut-off instead
)
fit <- epidist(as_epidist_marginal_model(ld),
               chains = 2, cores = 2, backend = "cmdstanr")
pred <- predict_delay_parameters(fit)   # posterior draws of meanlog and sdlog
c(meanlog = mean(pred$mu), sdlog = mean(pred$sigma))  # ~ 2.37, 0.54 (mean ~ 12.4 d)
```

Turn the posterior into a `delay` for `fit_cfr()`, using the posterior sd of each
parameter as the prior width (or fix it at the posterior mean for the
Ghani/Nishiura fit):

```{r use-estimate, eval = FALSE}
delay <- dist.spec::LogNormal(
  meanlog = dist.spec::Normal(mean(pred$mu), sd(pred$mu)),
  sdlog = dist.spec::Normal(mean(pred$sigma), sd(pred$sigma))
)
fit_cfr(d, delay = delay)
```

For a real-time estimate, set `obs_time` to the cut-off so epidist corrects for
the right-truncation. And if you estimate from the same deaths you then fit,
remember that `fit_cfr()` already co-estimates the delay jointly with `cfr` when
you pass `Normal()` priors; the separate fit is mainly for a delay taken from
historical or external data (say a previous outbreak), used as an informed prior
or a fixed delay.

## Using recovery timing

If your line list records when non-fatal cases recovered, add a `recovery_date`
column and `prepare_cfr_data()` times those recoveries. Here we simulate
recoveries that resolve more slowly than deaths (mean 21 vs 12.75 days):

```{r recovery}
set.seed(2)
llr <- simulate_linelist(n = 500, cfr = 0.55, onset_days = 45,
                         delay = dist.spec::Gamma(mean = 12.75, sd = 7),
                         recovery = dist.spec::Gamma(mean = 21, sd = 9))
dr <- prepare_cfr_data(llr, obs_time = max(llr$onset_date) - 3)
c(deaths = dr$n_death, recoveries = dr$n_recovery, censored = dr$n_cens)
```

By default the recovery timing is ignored: a recovered case contributes only the
cure factor `1 - cfr`. Pass a `recovery_delay` to fit the competing-risks model,
which uses *when* each case recovered: a recovered case contributes
`(1 - cfr) * f_R(r)`, and an unresolved case at time `t` contributes
`cfr * (1 - F_D(t)) + (1 - cfr) * (1 - F_R(t))`:

```{r recovery-fit, eval = FALSE}
onset_to_recovery <- dist.spec::LogNormal(
  meanlog = dist.spec::Normal(2.9, 0.3),
  sdlog = dist.spec::Normal(0.5, 0.2)
)
fit <- fit_cfr(dr, delay = onset_to_death, recovery_delay = onset_to_recovery)
summarise_cfr(fit)   # also reports recovery_mean / recovery_sd
```

Using recovery timing sharpens `cfr`: a case still unresolved long after onset is
unlikely to be a survivor, since it would probably have recovered by now, so the
model can tell it apart from a fatal case that has not yet died.

## Assumptions and caveats

The correction fixes a *timing* bias; it cannot repair the data.

- **Complete death ascertainment.** A death missing from the line list (e.g. a
  community death outside a treatment centre) is treated as a survivor and
  biases the CFR down. This is the dominant risk when ascertainment is
  ETC-centred.
- **Use the death *notification* date.** cfrnow counts a case as a death once
  its `death_date` is on or before the cut-off, so `death_date` should be the
  date the death entered the data, not the true date of death. If notification is
  prompt the two coincide. If it lags, the notification date lets the
  right-censoring absorb the delay (which then reads as onset-to-notification)
  and `cfr` stays correct, while the true date of death biases the real-time
  `cfr` down, since deaths not yet notified read as unresolved. If you only have
  the true date of death, instead set `obs_time` earlier, to a date by which
  reporting is near-complete (trading recency for completeness). A lag in
  notifying an *onset* is different: it only makes the case absent, which does
  not bias `cfr` unless notification depends on outcome.
- **Weak identifiability early on.** When few deaths have resolved, the CFR
  posterior stays close to its prior; `summarise_cfr()` flags this via
  `cfr_low_information`. Treat a flagged estimate as prior-driven.
- **Stationary delay and homogeneous CFR.** A pooled CFR lags reality as
  treatment access scales up; time-varying CFR is future work.
- **Delay family is chosen, not estimated.** Refit with a
  `dist.spec::Gamma()` delay to check sensitivity to the tail.
