---
title: "Real-time CFR with cfrnow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Real-time CFR with cfrnow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## The problem

The naive case fatality ratio, `deaths / cases`, is biased downward in real
time: recent cases have not yet had time to die. Restricting to cases with a
*resolved* outcome, `deaths / (deaths + recoveries)`, swaps that for an
upward bias, because deaths resolve faster than recoveries, so at any
mid-outbreak cut-off the resolved set is enriched for deaths.

`cfrnow` sidesteps both by never conditioning on resolution. Each case is fatal
with probability `cfr`; a fatal case dies at an interval-censored onset-to-death
delay `F`; every case still unresolved at the cut-off is right-censored,
contributing the mixture-cure survival term `1 - cfr * F(t)`. Because unresolved
cases are censored rather than dropped, the `deaths / (deaths + recoveries)`
enrichment cannot arise. Recovery dates aren't required, but when recorded a
recovered case is counted as a resolved non-fatal, which tightens the estimate.

```{r setup}
library(cfrnow)
```

## Choosing the model

Which model `fit_cfr()` runs is set by the data you have:

- With onset and death dates only, you get the default death-only fit:
  unresolved cases are censored and recoveries are left unmodelled.
- If you also have reliable recovery dates, pass a `recovery_delay` and the fit
  uses their timing (the competing-risks model).
- If you already trust an onset-to-death delay from elsewhere and the data are
  thin, fix that delay so only `cfr` is estimated (the Ghani/Nishiura estimator).

`obs_time` is a separate choice: a date gives a real-time cut-off; `NULL` gives a
retrospective fit, where every case has resolved.

## A line list

Your line list needs an `onset_date` and a `death_date` (`NA` for cases that
have not died); an optional `onset_lower`/`onset_upper` window widens the onset
censoring, and an optional `recovery_date` marks a non-fatal recovery so that
case is treated as resolved rather than censored. Here we simulate one with a
known CFR of 0.55 sampled part-way through a growing epidemic.

```{r simulate}
set.seed(1)
ll <- simulate_linelist(n = 500, cfr = 0.55, delay_mean = 12.75, delay_sd = 7,
                        delay_family = "gamma", onset_days = 45)
head(ll)
```

## Preparing the data at a cut-off

`prepare_cfr_data()` classifies each case at the observation cut-off into an
observed death, a right-censored survivor, or (retrospectively) a fully-resolved
non-death.

```{r prepare}
cutoff <- max(ll$onset_date) - 3
d <- prepare_cfr_data(ll, obs_time = cutoff)
c(cases = d$n_cases, deaths = d$n_deaths, censored = d$n_cens)
naive <- d$n_deaths / d$n_cases
round(naive, 3)  # the downward-biased naive ratio
```

## Fitting

`fit_cfr()` fits the mixture-cure model with cmdstanr. You supply the
onset-to-death delay as a [dist.spec](https://epiforecasts.io/dist.spec/)
distribution; there is no default, because a sensible delay depends on the
pathogen and setting. Give a native parameter a `Normal()` prior to co-estimate
it, or a fixed number / `Fixed()` to hold it fixed (fixing the whole delay gives
the Ghani/Nishiura estimator). Here we use the BDBV/Isiro onset-to-death prior.
This step needs a working CmdStan install and is not run when the vignette is
built.

```{r fit, eval = FALSE}
onset_to_death <- dist.spec::LogNormal(
  meanlog = dist.spec::Normal(2.41, 0.2),
  sdlog = dist.spec::Normal(0.51, 0.15)
)
fit <- fit_cfr(d, delay = onset_to_death)
summarise_cfr(fit)
#>     quantity   mean   q2.5    q50  q97.5  rhat ess_bulk
#> 1        cfr  0.579  0.528  0.579  0.625 1.001     3200
#> 2 delay_mean 12.700 11.500 12.700 13.900 1.000     3400
#> 3   delay_sd  7.200  6.300  7.200  8.200 1.000     3100
```

The corrected CFR (~0.58) recovers the true 0.55 and sits well above the naive
`r round(naive, 2)`, which is depressed by the many recent, not-yet-resolved
cases. The summary also carries `rhat`/`ess_bulk` and a `cfr_low_information`
attribute (see below). Pass `obs_time = NULL` to `prepare_cfr_data()` for a
retrospective, fully-resolved fit, which returns the naive ratio with the delay
as a nuisance.

Change the family or fix the delay by passing your own dist.spec distribution:

```{r delay-variants, eval = FALSE}
library(dist.spec)
# a gamma delay
fit_cfr(d, delay = Gamma(shape = Normal(3.3, 1), rate = Normal(0.26, 0.08)))
# a fixed delay: the Ghani/Nishiura fixed-F estimator (only cfr is estimated)
fit_cfr(d, delay = LogNormal(meanlog = 2.41, sdlog = 0.51))
```

Because the model is parameterised by the native parameters and reports
`delay_mean`/`delay_sd` as generated quantities, the summary looks the same
whichever family you choose; refit with a different family to check tail
sensitivity.

## Estimating the delay from data

Rather than hand-picking the prior, you can estimate the onset-to-death delay
from a line list with
[primarycensored](https://primarycensored.epinowcast.org/)'s
`fitdistdoublecens()`, which handles the double interval censoring (onset and
death each known only to the day) and, in real time, the right-truncation:

```{r estimate-delay, eval = FALSE}
library(primarycensored)   # also needs the fitdistrplus package

deaths <- ll[!is.na(ll$death_date), ]
delays <- as.numeric(deaths$death_date - deaths$onset_date)
cens <- data.frame(left = delays, right = delays + 1,  # death known to the day
                   pwindow = 1,                          # onset known to the day
                   D = Inf)                              # no right truncation
est <- fitdistdoublecens(cens, distr = "lnorm",
                         start = list(meanlog = 2, sdlog = 1))
coef(est)   # meanlog ~ 2.40, sdlog ~ 0.54  (mean ~ 12.8 days)
```

Turn the estimate into a `delay` for `fit_cfr()`, using the standard errors as
the prior widths (or fix it at the point estimate for the Ghani/Nishiura fit):

```{r use-estimate, eval = FALSE}
p <- coef(est)
se <- sqrt(diag(vcov(est)))
delay <- dist.spec::LogNormal(
  meanlog = dist.spec::Normal(p[["meanlog"]], se[["meanlog"]]),
  sdlog = dist.spec::Normal(p[["sdlog"]], se[["sdlog"]])
)
fit_cfr(d, delay = delay)
```

Two caveats. In real time, replace `D = Inf` with each case's follow-up
(`obs_time - onset`) so the delay is corrected for right-truncation. And if you
estimate from the same deaths you then fit, remember that `fit_cfr()` already
co-estimates the delay jointly with `cfr` when you pass `Normal()` priors; the
separate fit earns its keep mainly for a delay taken from historical or external
data (say a previous outbreak), used as an informed prior or a fixed delay.

## Using recovery timing

If your line list records when non-fatal cases recovered, add a `recovery_date`
column and `prepare_cfr_data()` times those recoveries. Here we simulate
recoveries that resolve more slowly than deaths (mean 21 vs 12.75 days):

```{r recovery}
set.seed(2)
llr <- simulate_linelist(n = 500, cfr = 0.55, recovery = TRUE,
                         delay_mean = 12.75, delay_sd = 7,
                         recovery_mean = 21, recovery_sd = 9, onset_days = 45)
dr <- prepare_cfr_data(llr, obs_time = max(llr$onset_date) - 3)
c(deaths = dr$n_death, recoveries = dr$n_recovery, censored = dr$n_cens)
```

By default the recovery timing is ignored: a recovered case contributes only the
cure factor `1 - cfr`. Pass a `recovery_delay` to fit the competing-risks model,
which uses *when* each case recovered: a recovered case contributes
`(1 - cfr) * f_R(r)`, and an unresolved case at time `t` contributes
`cfr * (1 - F_D(t)) + (1 - cfr) * (1 - F_R(t))`:

```{r recovery-fit, eval = FALSE}
onset_to_recovery <- dist.spec::LogNormal(
  meanlog = dist.spec::Normal(2.9, 0.3),
  sdlog = dist.spec::Normal(0.5, 0.2)
)
fit <- fit_cfr(dr, delay = onset_to_death, recovery_delay = onset_to_recovery)
summarise_cfr(fit)   # also reports recovery_mean / recovery_sd
```

Using recovery timing sharpens `cfr`: a case still unresolved long after onset is
unlikely to be a survivor, since it would probably have recovered by now, so the
model can tell it apart from a fatal case that has not yet died.

## Assumptions and caveats

The correction fixes a *timing* bias; it cannot repair the data.

- **Complete death ascertainment.** A death missing from the line list (e.g. a
  community death outside a treatment centre) is treated as a survivor and
  biases the CFR down. This is the dominant risk when ascertainment is
  ETC-centred.
- **Deaths reported without lag.** cfrnow assumes a death is in the data by the
  cut-off. If reporting lags, recent deaths are missing and read as still
  unresolved, so the real-time `cfr` comes out too low and revises upward as
  deaths arrive; treat the latest weeks as provisional. If the line list also
  records *when* each death was reported (not just when it occurred), check the
  lag; if it is material, move `obs_time` back to a date by which reporting is
  near-complete (`prepare_cfr_data()` then censors everyone there, trading
  recency for completeness). Folding a death-reporting delay into the likelihood
  is future work.
- **Weak identifiability early on.** When few deaths have resolved, the CFR
  posterior stays close to its prior; `summarise_cfr()` flags this via
  `cfr_low_information`. Treat a flagged estimate as prior-driven.
- **Stationary delay and homogeneous CFR.** A pooled CFR lags reality as
  treatment access scales up; time-varying CFR is future work.
- **Delay family is chosen, not estimated.** Refit with a
  `dist.spec::Gamma()` delay to check sensitivity to the tail.
