---
title: "Real-time CFR with cfrnow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Real-time CFR with cfrnow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## The problem

The naive case fatality ratio, `deaths / cases`, is biased downward in real
time: recent cases have not yet had time to die. Restricting to cases with a
*resolved* outcome, `deaths / (deaths + recoveries)`, swaps that for an
upward bias, because deaths resolve faster than recoveries, so at any
mid-outbreak cut-off the resolved set is enriched for deaths.

`cfrnow` sidesteps both by never conditioning on resolution. Each case is fatal
with probability `cfr`; a fatal case dies at an interval-censored onset-to-death
delay `F`; every case still unresolved at the cut-off is right-censored,
contributing the mixture-cure survival term `1 - cfr * F(t)`. Because unresolved
cases are censored rather than dropped, the `deaths / (deaths + recoveries)`
enrichment cannot arise. Recovery dates aren't required, but when recorded a
recovered case is counted as a resolved non-fatal, which tightens the estimate.

```{r setup}
library(cfrnow)
```

## A line list

Your line list needs an `onset_date` and a `death_date` (`NA` for cases that
have not died); an optional `onset_lower`/`onset_upper` window widens the onset
censoring, and an optional `recovery_date` marks a non-fatal recovery so that
case is treated as resolved rather than censored. Here we simulate one with a
known CFR of 0.55 sampled part-way through a growing epidemic.

```{r simulate}
set.seed(1)
ll <- simulate_linelist(n = 500, cfr = 0.55, delay_mean = 12.75, delay_sd = 7,
                        delay_family = "gamma", onset_days = 45)
head(ll)
```

## Preparing the data at a cut-off

`prepare_cfr_data()` classifies each case at the observation cut-off into an
observed death, a right-censored survivor, or (retrospectively) a fully-resolved
non-death.

```{r prepare}
cutoff <- max(ll$onset_date) - 3
d <- prepare_cfr_data(ll, obs_time = cutoff)
c(cases = d$n_cases, deaths = d$n_deaths, censored = d$n_cens)
naive <- d$n_deaths / d$n_cases
round(naive, 3)  # the downward-biased naive ratio
```

## Fitting

`fit_cfr()` fits the mixture-cure model with cmdstanr. The onset-to-death delay
is supplied as a [dist.spec](https://epiforecasts.io/dist.spec/) distribution.
`default_delay()` is a lognormal with `Normal()` priors on its native parameters
(the Julia reference values); give a parameter as a number or `Fixed()` to hold
it fixed, and fixing the whole delay gives the Ghani/Nishiura fixed-delay
estimator. This step needs a working CmdStan install and is not run when the
vignette is built.

```{r fit, eval = FALSE}
fit <- fit_cfr(d)                       # default_delay(): estimated lognormal
summarise_cfr(fit)
#>     quantity   mean   q2.5    q50  q97.5  rhat ess_bulk
#> 1        cfr  0.579  0.528  0.579  0.625 1.001     3200
#> 2 delay_mean 12.700 11.500 12.700 13.900 1.000     3400
#> 3   delay_sd  7.200  6.300  7.200  8.200 1.000     3100
```

The corrected CFR (~0.58) recovers the true 0.55 and sits well above the naive
`r round(naive, 2)`, which is depressed by the many recent, not-yet-resolved
cases. The summary also carries `rhat`/`ess_bulk` and a `cfr_low_information`
attribute (see below). Pass `obs_time = NULL` to `prepare_cfr_data()` for a
retrospective, fully-resolved fit, which returns the naive ratio with the delay
as a nuisance.

## Using recovery dates

If your line list records when non-fatal cases recovered, add a `recovery_date`
column. A case recovered by the cut-off is then counted as a resolved non-death
rather than right-censored, so the fit uses more information. Here we simulate
recoveries (which resolve more slowly than deaths, mean 21 vs 12.75 days):

```{r recovery}
set.seed(2)
llr <- simulate_linelist(n = 500, cfr = 0.55, recovery = TRUE,
                         delay_mean = 12.75, delay_sd = 7,
                         recovery_mean = 21, recovery_sd = 9, onset_days = 45)
dr <- prepare_cfr_data(llr, obs_time = max(llr$onset_date) - 3)
c(deaths = dr$n_deaths, resolved = dr$n_resolved, censored = dr$n_cens)
```

Without the `recovery_date` column those recovered cases would all be censored
(`resolved = 0`). Counting them as resolved sharpens the posterior for `cfr`,
because a recovered case is known to be non-fatal rather than possibly
fatal-but-not-yet-resolved.

## Assumptions and caveats

The correction fixes a *timing* bias; it cannot repair the data.

- **Complete death ascertainment.** A death missing from the line list (e.g. a
  community death outside a treatment centre) is treated as a survivor and
  biases the CFR down. This is the dominant risk when ascertainment is
  ETC-centred.
- **Deaths known on the day they occur.** Notification lag reintroduces the bias
  the model removes, via the data rather than the delay; nowcast or caveat the
  death series if that lag is non-trivial.
- **Weak identifiability early on.** When few deaths have resolved, the CFR
  posterior stays close to its prior; `summarise_cfr()` flags this via
  `cfr_low_information`. Treat a flagged estimate as prior-driven.
- **Stationary delay and homogeneous CFR.** A pooled CFR lags reality as
  treatment access scales up; time-varying CFR is future work.
- **Delay family is chosen, not estimated.** Refit with a
  `dist.spec::Gamma()` delay to check sensitivity to the tail.
