---
title: "Real-time CFR with cfrnow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Real-time CFR with cfrnow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## The problem

The naive case fatality ratio, `deaths / cases`, is biased downward in real
time: recent cases have not yet had time to die. Restricting to cases with a
*resolved* outcome, `deaths / (deaths + recoveries)`, swaps that for an upward
bias, because deaths resolve faster than recoveries.

`cfrnow` sidesteps both by never conditioning on resolution. Each case is fatal
with probability `cfr`; a fatal case dies at an interval-censored onset-to-death
delay `F`; every case still unresolved at the cut-off is right-censored,
contributing the mixture-cure survival term `1 - cfr * F(t)`. With `F` fixed this
is the Ghani/Nishiura estimator; here `F` is co-estimated and its uncertainty
propagated.

`cfrnow` is registered as an [epidist](https://epidist.epinowcast.org/) model
type, so `cfr` and the delay both take `brms` formulas — that is how you put
covariates, or a smooth time effect, on the CFR.

```{r setup}
library(cfrnow)
```

## A line list

Your line list needs an `onset_date` and a `death_date` (`NA` for cases that
have not died); an optional `onset_lower`/`onset_upper` window widens the onset
censoring. Here we simulate one with a known CFR of 0.55 sampled part-way
through a growing epidemic.

```{r simulate}
set.seed(1)
ll <- simulate_linelist(n = 500, cfr = 0.55, onset_days = 45,
                        delay = LogNormal(mean = 12.75, sd = 7))
head(ll)
```

## Preparing the data at a cut-off

`prepare_cfr_data()` classifies each case at the observation cut-off into an
observed death, a right-censored survivor, or (retrospectively) a fully-resolved
non-death. `obs_time = NULL` gives a retrospective fit where every case has
resolved.

```{r prepare}
cutoff <- max(ll$onset_date) - 3
d <- prepare_cfr_data(ll, obs_time = cutoff)
c(cases = d$n_cases, deaths = d$n_deaths, censored = d$n_cens)
naive <- d$n_deaths / d$n_cases
round(naive, 3)  # the downward-biased naive ratio
```

## Fitting

`fit_cfr()` is a thin wrapper over `epidist::epidist()`. With no formula it fits a
single CFR and co-estimates the onset-to-death delay; the default prior is a
weakly-informative logit-normal on the CFR. `summary()` reports the corrected CFR
and the delay moments (days) with convergence diagnostics.

```{r fit, eval = FALSE}
fit <- fit_cfr(d, backend = "cmdstanr")
summary(fit)
```

The corrected `cfr` sits above the naive ratio and, with enough resolved deaths,
covers the true 0.55. `summary()` also carries a `cfr_low_information` attribute,
`TRUE` when the CFR posterior has barely moved from its prior — expected very
early in an outbreak, when the estimate is prior-driven.

## Covariates and time-varying CFR

Because the model is an `epidist` subclass, a `brms::bf()` puts a formula on `cfr`
(or on the delay location `mu`). This is how you stratify or, with a smooth on
time, let the CFR vary:

```{r formula, eval = FALSE}
fit_cfr(d, formula = brms::bf(mu ~ 1, cfr ~ age_group))  # CFR by covariate
fit_cfr(d, formula = brms::bf(mu ~ 1, cfr ~ s(week)))    # time-varying CFR
```

## Delay family and fixing the delay

The delay `family` is `lognormal()` (default) or `Gamma()`; they differ in tail
weight, which sets how quickly a recent case counts as "probably cured". The
family is not identifiable from sparse data, so check sensitivity by refitting
the other. To hold the delay fixed (the Ghani/Nishiura estimator) supply a tight
prior on `mu`/`sigma` via the `prior` argument.

```{r family, eval = FALSE}
fit_cfr(d, family = stats::Gamma(link = "log"))          # gamma delay
```

## Recovery timing

If the line list has a `recovery_date` column, `prepare_cfr_data()` times those
recoveries and `fit_cfr()` switches to a two-outcome mixture-cure model that also
estimates the onset-to-recovery delay (reported as `recovery_mean`/`recovery_sd`
by `summary()`). This sharpens `cfr` when discharge recording is reliable, but an
actually-recovered case whose recovery is *unrecorded* stays censored and is
pushed toward the fatal branch over time, biasing `cfr` up — so prefer the
death-only default where discharge data are patchy.

## Caveats

The correction fixes a *timing* bias, not the data. A death that never reaches
the line list is treated as a survivor and biases `cfr` down. Use the death
*notification* date in `death_date` so the censoring absorbs any reporting lag;
the delay the model then estimates is onset-to-notification.
