---
title: "Real-time CFR with curecfr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Real-time CFR with curecfr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## The problem

The naive case fatality ratio, `deaths / cases`, is biased **downward** in real
time: recent cases have not yet had time to die. Restricting to cases with a
*resolved* outcome, `deaths / (deaths + recoveries)`, swaps that for an
**upward** bias, because deaths resolve faster than recoveries, so at any
mid-outbreak cut-off the resolved set is enriched for deaths.

`curecfr` sidesteps both by never conditioning on resolution. Each case is fatal
with probability `cfr`; a fatal case dies at an interval-censored onset-to-death
delay `F`; every case still alive at the cut-off is right-censored, contributing
the mixture-cure survival term `1 - cfr * F(t)`. Recovery times are not used, so
the resolved-set enrichment cannot arise.

```{r setup}
library(curecfr)
```

## A line list

Your line list needs an `onset_date` and a `death_date` (`NA` for cases that
have not died); an optional `onset_lower`/`onset_upper` window widens the onset
censoring. Here we simulate one with a known CFR of 0.55 sampled part-way
through a growing epidemic.

```{r simulate}
set.seed(1)
ll <- simulate_linelist(n = 500, cfr = 0.55, delay_mean = 12.75, delay_sd = 7,
                        delay_family = "gamma", onset_days = 45)
head(ll)
```

## Preparing the data at a cut-off

`prepare_cfr_data()` classifies each case at the observation cut-off into an
observed death, a right-censored survivor, or (retrospectively) a fully-resolved
non-death.

```{r prepare}
cutoff <- max(ll$onset_date) - 3
d <- prepare_cfr_data(ll, obs_time = cutoff)
c(cases = d$n_cases, deaths = d$n_deaths, censored = d$n_cens)
naive <- d$n_deaths / d$n_cases
round(naive, 3)  # the downward-biased naive ratio
```

## Fitting

`fit_cfr()` fits the mixture-cure model with cmdstanr. The delay is
parameterised by its mean and standard deviation, so the choice of family
(`"gamma"` by default, or `"lognormal"`) leaves the priors and summaries
unchanged. This step needs a working CmdStan install and is not run when the
vignette is built.

```{r fit, eval = FALSE}
fit <- fit_cfr(d, delay_family = "gamma")
summarise_cfr(fit)
#>     quantity   mean   q2.5    q50  q97.5
#> 1        cfr  0.579  0.528  0.579  0.625
#> 2 delay_mean 12.700 11.500 12.700 13.900
#> 3   delay_sd  7.200  6.300  7.200  8.200
```

The corrected CFR (~0.58) recovers the true 0.55 and sits well above the naive
`r round(naive, 2)`, which is depressed by the many recent, not-yet-resolved
cases. Pass `obs_time = NULL` to `prepare_cfr_data()` for a retrospective,
fully-resolved fit, which returns the naive ratio with the delay as a nuisance.

## Notes

- `curecfr_families()` lists the supported onset-to-death delay families.
- `curecfr_priors()` sets the priors on the delay mean/sd (days) and the CFR.
- The mixture-cure correction fixes the *timing* bias only. It still assumes
  deaths are ascertained; under-counted deaths (for example community deaths
  outside treatment centres) are a separate problem it cannot repair.
