---
title: "Stratified and partially-pooled CFR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Stratified and partially-pooled CFR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Why stratify, and why pool

CFR usually varies by group -- age, treatment centre, calendar period -- and you
often want a per-group number. Two naive options both disappoint when some
groups are sparse. Fitting each group independently returns noisy estimates
wherever the data are thin. Ignoring the grouping and reporting one CFR hides
real heterogeneity.

*Partial pooling* is the middle path: fit a single model in which each group's
CFR is drawn from a common distribution, so data-rich groups are estimated
almost independently while data-poor groups are shrunk toward the shared mean.
Because `cfrnow` is an [epidist](https://epidist.epinowcast.org/) model, the CFR
takes a `brms` formula, so pooling is a one-line change: `cfr ~ (1 | group)`.

## A six-site line list

We simulate six treatment centres with a wide range of sample sizes (18 to 400)
and true CFRs scattered around a common mean. The data-rich sites anchor the
mean; the sparse ones are where pooling earns its keep. Site labels are
synthetic.

```{r data}
library(cfrnow)
set.seed(20260714)

sites <- data.frame(
  site = c("A", "B", "C", "D", "E", "F"),
  n    = c(400, 250, 120, 60, 30, 18),
  cfr  = c(0.50, 0.42, 0.55, 0.30, 0.62, 0.40)
)

death_delay    <- LogNormal(mean = 11, sd = 6.5)
recovery_delay <- LogNormal(mean = 16, sd = 8)

ll <- do.call(rbind, Map(function(s, n, cfr) {
  cbind(
    simulate_linelist(
      n = n, cfr = cfr, delay = death_delay,
      recovery = recovery_delay, onset_days = 45
    ),
    site = s
  )
}, sites$site, sites$n, sites$cfr))

d <- prepare_cfr_data(ll, obs_time = as.Date("2026-02-20"), covariates = "site")
c(cases = d$n_cases, deaths = d$n_deaths)
```

## Three fits: complete, no, and partial pooling

The only thing that changes across the three fits is the CFR formula; the delay
and recovery delay are co-estimated in every case. (Two short chains keep the
vignette quick; use more for real work.)

```{r fits, results = "hide", message = FALSE, warning = FALSE}
otd <- LogNormal(meanlog = Normal(2.4, 0.2), sdlog = Normal(0.5, 0.15))
otr <- LogNormal(meanlog = Normal(2.7, 0.2), sdlog = Normal(0.5, 0.15))
args <- list(
  delay = otd, recovery_delay = otr, cfr_prior = Beta(1, 1),
  backend = "cmdstanr", chains = 2, iter = 1000, refresh = 0, seed = 1
)

# complete pooling: a single CFR
fit_pool    <- do.call(fit_cfr, c(list(d, formula = brms::bf(mu ~ 1, cfr ~ 1)), args))

# no pooling: an independent CFR per site (one estimated logit-CFR per site)
fit_sep     <- do.call(fit_cfr, c(list(d, formula = brms::bf(mu ~ 1, cfr ~ 0 + site)), args))

# partial pooling: per-site CFR shrunk to a common mean
fit_partial <- do.call(fit_cfr, c(list(d, formula = brms::bf(mu ~ 1, cfr ~ (1 | site))), args))
```

## Reading the estimates out

`summary()` reports the CFR per group: a single `cfr` row for the complete-pool
fit, and one `cfr[<site>]` row per site for a grouped fit.

```{r summary-grouped}
summary(fit_sep)
```

Collect those per-site CFR rows from the two grouped fits for the picture below;
the complete-pool fit supplies the shared mean.

```{r extract}
cfr_rows <- function(fit, approach) {
  s <- summary(fit)
  r <- s[grepl("^cfr\\[", s$quantity), ] # the cfr[<site>] rows
  data.frame(
    approach = approach,
    site = sub("^cfr\\[(.*)\\]$", "\\1", r$quantity),
    median = r$q50, lower = r$q2.5, upper = r$q97.5
  )
}

est <- rbind(cfr_rows(fit_sep, "no pool"), cfr_rows(fit_partial, "partial pool"))
est <- merge(est, sites[, c("site", "n", "cfr")], by = "site")
names(est)[names(est) == "cfr"] <- "truth"

s_pool <- summary(fit_pool)
pool_mean <- s_pool$q50[s_pool$quantity == "cfr"]
```

## Shrinkage in one picture

Each site shows its no-pool and partial-pool estimate against the truth (×) and
the complete-pool mean (dotted). The data-rich sites (A, B) barely move; the
sparse ones (E, F) are pulled toward the mean and tightened -- trading a little
bias for a large drop in variance.

```{r shrinkage, fig.width = 6.5, fig.height = 3.6}
library(ggplot2)
est$approach <- factor(est$approach, levels = c("no pool", "partial pool"))
est$site <- factor(est$site, levels = rev(sites$site))

ggplot(est, aes(median, site, colour = approach)) +
  geom_vline(xintercept = pool_mean, linetype = 3) +
  geom_point(aes(x = truth), shape = 4, size = 2.4, colour = "grey30") +
  geom_pointrange(aes(xmin = lower, xmax = upper), position = position_dodge(width = 0.55)) +
  scale_y_discrete(labels = function(s) paste0(s, "  (n=", sites$n[match(s, sites$site)], ")")) +
  labs(
    x = "CFR: median & 95% CrI   (× = truth, dotted = complete-pool mean)",
    y = NULL, colour = NULL
  ) +
  theme_minimal()
```

## Let the delay vary too

An "irregular" pooled delay is often just several populations stacked. If sites
resolve on different timescales, pool the delay by site as well, so each keeps
its own onset-to-death distribution and the shared shape stops being multimodal:

```{r delay-pool, eval = FALSE}
fit_both <- do.call(fit_cfr, c(
  list(d, formula = brms::bf(mu ~ (1 | site), cfr ~ (1 | site))), args
))
```

## Check the fit

Confirm the parametric delay reproduces the observed delays before trusting the
tail extrapolation the real-time correction relies on:

```{r ppcheck, fig.width = 6.5, fig.height = 3.6, message = FALSE, warning = FALSE}
pp_check_cfr(fit_partial, type = "delay")
```

## Takeaways

- Pooling is a formula change: `cfr ~ (1 | group)` co-fits every group in one
  model and shrinks the sparse ones toward a shared mean.
- `summary()` reports one `cfr[<group>]` row per group for a grouped fit, so the
  per-group CFRs come straight out of it.
- If the pooled onset-to-death delay looks irregular, suspect a mixture: pool the
  delay by group too before reaching for a more flexible family.
